Outliers Detection and Comparison of Origin-Destination Flows with Data Depth

### 2018-10-09 giscience_1/giscience2018_new.tex
xx

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author[1]{Myeong-Hun Jeong}
\author[2]{Junjun Yin}
\author[3]{Shaowen Wang}
\affil[1]{Department of Civil Engineering,  Gwangju, Republic of Korea\\
  \texttt{mhjeong@chosun.ac.kr}}
\affil[2]{Social Science Research Institute, Penn State University, PA, USA\\
  \texttt{jyin@psu.edu}}
\affil[3]{Departmet of Geography and Geographic Information Science, the University of Illinois at Urbana-Champaign, IL, USA\\
	\texttt{shaowen@illinois.edu}}
\authorrunning{M.-H. Jeong, J. Yin and S. Wang} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

@@

\section{Introduction}

With the rapid rise in ubiquity of geo-tagged sensors, knowledge discovery power is enhanced by extracting interesting patterns from spatiotemporal big data in various domains. In particular, the location-acquisition technologies generate large volumes of movement data such as people, animals, vehicles and natural phenomena. These big movement data help us understand moving objects, and discover the hidden patterns. Thus, trajectory mining leads to solutions for important research problems in different fields such as urban planning \cite{mazimpaka15AGILE}, transportation \cite{chen13Percom}, environment \cite{devarakonda13SIGKDD}, and public security and safety \cite{buchin14JOSIS}.

This paper presents a new algorithm which not only estimates origination-destination (OD)'s flows anomaly but also conducts hypothesis testing between two different OD flows. This paper utilizes a New York City Taxi data set that has the origin and destination of each movement without the whole actual trajectory routes. The analysis of digital footprints obtained from taxi data has a crucial role in  understanding crowd patterns and planning urban and transportation investments in administrative authorities.

In recent years, researchers have investigated a variety of approaches to trajectory data mining. There are generally four categories that classify trajectory mining methods: clustering, classification, frequent/group pattern mining, and outlier detection \cite{mazimpaka16JOSIS,zheng15ACMTIST}. These techniques can be used independently or combinatorially for trajectory mining application problems.

In particular, this study focuses on outlier detection of OD flows. Outlier detection aims to find trajectories that do not follow the typical flow of trajectory data sets for characterizing the connectivity between regions \cite{mazimpaka16JOSIS}. \cite{fontes13GeoInfo,lee08ICDE} use Euclidean distance to find outlier patterns from trajectories.  \cite{pan13ACMGIS,liu12IJGIS} raise the questions of Euclidean distance approach due to the loss of local features and unavailability when external factors affect the trajectories (e.g., topography, land cover or weather condition).  \cite{pan13ACMGIS,liu12IJGIS} addresses this by using robust distance measurements (i.e., Mahalanobis distance \cite{pan13ACMGIS} and relative distance  \cite{liu12IJGIS}). Further, structural features  \cite{yuan11JCIS} and a data-induced random tree  \cite{zhang11UC} are exploited to detect anomalous trajectories instead of using distance or density. In a similar fashion, this study considers the center regions of OD flows with data depth in order to robustly detect the trajectory outliers.

In addition, visual analytics are a common approach to analyze OD flow data. Visual representation of massive movement data enables comprehensive exploration of data and results in understanding complex flow trends. Aggregation and generalization of movement data are frequently utilized \cite{andrienko08VAST,adrienko11IEEETVCG,guo14IEEETVCG}. While visual analytics can help to extract inherent patterns from massive data, it is difficult to compare two different OD flows based on a hypothesis testing. In other words, it is complicated to comprehend how two OD flows differ and by how much. This paper uses bivariate hypothesis testing methods based on data depth to understand how different two OD flow data sets are in terms of the amount of spatial extent and a shift in location.

Further, flow mapping approaches frequently suffer from the modifiable area unit problem (MAUP). For instance, it is not guaranteed that different aggregations via location can present coherent patterns. Kernel-based flow estimation and smoothing are used to overcome different spatial resolution \cite{guo14IEEETVCG}. This study uses traffic analysis zones in New York City as a base unit. Thus, it does not consider MAUP problem. In addition, New York City Taxi data includes the origin and destination within traffic analysis zones, which ignores the actual trajectory route. It is not necessary to reconstruct individual movements for flow estimation (see \cite{duckham16ICGIS}).

In summary, this paper presents a new algorithm which conducts outlier detection as well as a hypothesis testing from OD flows data. Our approach investigates the central regions of  OD flows based on data depth to detect OD flows anomaly and conduct a hypothesis testing between two different OD flow data sets.

In the remainder of this paper, Section \ref{sec:methods} overviews how to detect OD flows outliers and conduct  a hypothesis testing between two different OD flows with the concept of data depth. Experimental design and the evaluation of the algorithm are presented in Section \ref{sec:experiments}. These results are discussed in Section \ref{sec:discussion}. Section 5 concludes with a summary and future perspectives.

@@

\subsection{Data Depth}
Data depth measures the centrality of a point with regard to a given data sets in $\mathbb{R}^d$.  The notion of data depth (i.e., Tukey's halfspace depth) is originally developed by \cite{tukey75ICM}, which generalizes the univariate concept of ranking to multivariate data. It presents how deeply a point is located within a given data sets by ordering their degree of centrality.

In general, the halfspace depth (HD) of a point $x$ in  $\mathbb{R}^d$ is defined as the minimum probability, $P$ on  $\mathbb{R}^d$, associated with any closed halfspace containing $x$ \cite{liu00AS}. 
\begin{equation*}\label{eq:hd}
HD(x;P) = inf\{P(H): \text{H a closed halfspace}, x \in H\}\text{, } x \in \mathbb{R}^d.
\end{equation*}

For the univariate case, all values less than or equal (greater than or equal) to $x$ form a closed halfspace. 
In Figure \ref{fig:hd_uni}, the probability of values less than or equal to 4 is  $2/7$ and the probability greater than or equal to 4 is $6/7$. The halfspace depth of 4 is $2/7$ which is the minimum probability carried by any closed halfspace containing 4. Further, 14 has the largest halfspace depth, which is the usual sample median. However, the polluted point inflates the standard error of the sample mean, thereby giving a distorted view of the data.

@@

In a similar fashion, the halfspace depth of $x$ for the bivariate case is defined as the minimal number of data points in any closed halfspace, determined by a hyperplane through $x$ \cite{rousseeuw96RSS}. For example, the line through $x$ is rotated by $180^{\circ}$ in Figure \ref{fig:hd_bi}. The halfspace depth of $x$ is determined by the smallest portion of data that are separated by such a hyperplane (e.g.,  the halfspace depth of $x$ is $3/13$, determined by the dotted line).

@@

\subsection{OD Trajectory Outlier Detection Based on Depth}
The center-outward ordering in data depth is closely related to the detection of outliers. The upper level sets of data depth in $\mathbb{R}^2$ form the central regions. The most central region can be regarded as a median. Conversely, the lower level sets of data depth, coincide with large distance from the center, can be regarded as outlyingness. \cite{rousseeuw99AS,aplpackR} utilizes this concept to generate a bag plot, which is analogous to the one-dimensional box plot based on data depth. This paper uses the bag plot in order to identify the outliers of OD flows. Before explaining the method of outliers detection, we define a couple of definitions.

@@

The bag plot is able to present the location (the depth median), spread (the size of bag), correlation (the orientation of the bag), and skewness (the shape of the bag and the fence) of data \cite{rousseeuw99AS}. For example, we can see which forward OD flows have higher counts compared with reverse OD flows; the relatively linear correlation between forward OD flows and reverse OD flows; and the skewness of forward (reverse) OD flows in Figure \ref{fig:OD_0521}.

In addition, it is possible to detect the outliers of OD flows of two different time stamps. In Figure \ref{fig:OD_0721_0521}, we visualize the OD flows of two different days. Thus, we can not only see where is the central region of OD flows, but also differentiate which OD flows are significantly different compared with two different OD flows.

Further, the OD flows of high active areas in a city are more likely to have large volume of trips. We use set operations to detect such outliers. For example, we regard OD flows on July 01  as control data set ($control$); OD flows on May 21 as test data set ($test$); and the combination of two OD flows as combination data set ($combination$) in Figure \ref{fig:bagplot_0521_0701}. Then we can calculate the intersection of three outliers sets ($control \cap test \cap combination$), which are represented as rectangle symbols in Figure \ref{fig:OD_outliers_high}.

@@

\subsection{OD Trajectory Comparisons Based on Depth}
Data depth can compare bivariate data related with two independent groups \cite{liu93JASA}. A t-test can be used to compare two means from independent groups. For example, the t-test reveals whether two OD flows' means  are different at two different time stamps. However, it is worth examining how groups differ in terms of scale. The comparisons of central regions in data depth compare the marginal distribution, thereby considering the overall structure of the data \cite{wilcox03MBR}.

>>>>

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author[1]{Myeong-Hun Jeong}
\author[2]{Junjun Yin}
\author[3]{Shaowen Wang}
\affil[1]{Department of Civil Engineering, Chosun University, Gwangju, Republic of Korea\\
  \texttt{mhjeong@chosun.ac.kr}}
\affil[2]{Social Science Research Institute; Institute for CyberScience, Penn State University, PA, USA\\
  \texttt{jyin@psu.edu}}
\affil[3]{Departmet of Geography and Geographic Information Science, University of Illinois at Urbana-Champaign, IL, USA\\
	\texttt{shaowen@illinois.edu}}
\authorrunning{M.-H. Jeong, J. Yin and S. Wang} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

@@

\section{Introduction}
With the rapid rise in ubiquity of geolocation-aware sensors, knowledge discovery is greatly enhanced by extracting and mining interesting patterns from spatiotemporal Big Data in various domains.
In particular, the location-acquisition technologies generate large volumes of movement data, which were used to track people, animals, vehicles, and even natural phenomena.
Such data help us better model moving objects and reveal hidden patterns that are important to urban planning and its applications in characterizing urban human mobility \cite{jiang2012discovering}, accessibility of the activity space \cite{kwan1998space}, and sustainability of the urban systems \cite{alberti2003integrating}. 
Importantly, trajectory mining leads to solutions for addressing important research problems in different fields, such as urban planning \cite{mazimpaka15AGILE}, transportation \cite{chen13Percom}, environment \cite{devarakonda13SIGKDD}, and public security and safety \cite{buchin14JOSIS}.
This paper presents a new algorithm which not only estimates origination-destination (OD)'s flows anomaly but also conducts hypothesis testing between two sets of different OD flows. The algorithm is applied on the New York City taxi data set, where each record contains the origin and destination of each trip (without intermediate locations of the actual routes).
We believe the our methods for analyzing taxi trip data has a promising application in understanding crowd patterns and planning urban and transportation investments by the administrative authorities.

In recent years, researchers have investigated a variety of approaches to trajectory data mining.
In general, the contemporary trajectory mining methods can be classified into four categories: clustering, classification, frequent/group pattern mining, and outlier detection \cite{mazimpaka16JOSIS,zheng15ACMTIST}. These techniques can be used independently or combinatorially for trajectory mining applications. For this study, we focus on outlier detection of OD flows. Outlier detection aims to find trajectories that do not follow the typical flow of trajectory data sets for characterizing the connectivity between regions \cite{mazimpaka16JOSIS}. \cite{fontes13GeoInfo,lee08ICDE} use Euclidean distance to find outlier patterns from trajectories. \cite{pan13ACMGIS,liu12IJGIS} raise the questions of Euclidean distance approach due to the loss of local features and unavailability when external factors affect the trajectories (e.g., topography, land cover or weather condition).  \cite{pan13ACMGIS,liu12IJGIS} addresses this by using robust distance measurements (i.e., Mahalanobis distance \cite{pan13ACMGIS} and relative distance \cite{liu12IJGIS}). Further, structural features \cite{yuan11JCIS} and a data-induced random tree  \cite{zhang11UC} are exploited to detect anomalous trajectories instead of using distance or density. In a similar fashion, this study considers the center regions of OD flows with data depth in order to robustly detect the trajectory outliers.

In addition, utilizing visual analytics is a common approach to analyze OD flow data.
Visual representation of massive movement data enables comprehensive exploration of data and results in understanding complex flow trends.
Aggregation and generalization of movement data are frequently utilized \cite{andrienko08VAST,adrienko11IEEETVCG,guo14IEEETVCG}.
While visual analytics can help to extract inherent patterns from massive data, it is difficult to quantatively compare two sets of different OD flows based on a hypothesis testing.
In other words, it is complicated to comprehend how two OD flows differ and, more importantly, by how much.
In this light, this paper uses bivariate hypothesis testing methods based on data depth to understand how different two OD flow data sets are in terms of the amount of spatial extent and a shift in location.

It is worthy noting that flow mapping approaches frequently suffer from the modifiable areal unit problem (MAUP). For instance, it is not guaranteed that different aggregations via location can present coherent patterns.
Kernel-based flow estimation and smoothing are used to overcome different spatial resolution \cite{guo14IEEETVCG}. 
Instead of focusing on finding the best areal unit to partioning the urban space and aggregating the OD flows, this study chooses traffic analysis zones in New York City as a base unit. However, the mothods can be adapted to other choices of areal units.
In addition, New York City taxi trip data include the origin and destination within traffic analysis zones, which ignores the intermediate locations of the actual routes.
It is not necessary to reconstruct individual movements for flow estimation (see \cite{duckham16ICGIS}).

In summary, this paper presents a new algorithm which conducts outlier detection as well as a hypothesis testing from OD flows data. Our approach investigates the central regions of OD flows based on data depth to detect OD flows anomaly and conduct a hypothesis testing between two different OD flow data sets. The remainder of this paper is organized as follows: Section \ref{sec:methods} overviews how to detect OD flows outliers and conduct  a hypothesis testing between two different OD flows with the concept of data depth. 
Experimental design and the evaluation of the algorithm are presented in Section \ref{sec:experiments}. 
These results are discussed in Section \ref{sec:discussion}. 
Section 5 concludes with a summary and future perspectives.

@@

\subsection{Data Depth}
Data depth measures the centrality of a point with regard to a given data sets in $\mathbb{R}^d$.
The notion of data depth (i.e., Tukey's halfspace depth) is originally developed by \cite{tukey75ICM}, which generalizes the univariate concept of ranking to multivariate data.
It reveals how deeply a point is located within a given data set by ordering their degree of centrality. 
In general, the halfspace depth (HD) of a point $x$ in  $\mathbb{R}^d$ is defined as the minimum probability, $P$ on  $\mathbb{R}^d$, associated with any closed halfspace containing $x$ \cite{liu00AS}. 
\begin{equation*}\label{eq:hd}
HD(x;P) = inf\{P(H): \text{H a closed halfspace}, x \in H\}\text{, } x \in \mathbb{R}^d.
\end{equation*}

For the univariate case, all values less than or equal (greater than or equal) to $x$ form a closed halfspace. 
In Figure \ref{fig:hd_uni}, the probability of values less than or equal to 4 is  $2/7$ and the probability greater than or equal to 4 is $6/7$. 
The halfspace depth of 4 is $2/7$, which is the minimum probability carried by any closed halfspace containing 4. 
Further, 14 has the largest halfspace depth, which is the usual sample median. 
However, the polluted point inflates the standard error of the sample mean, thereby giving a distorted view of the data.

@@

In a similar fashion, the halfspace depth of $x$ for the bivariate case is defined as the minimal number of data points in any closed halfspace, which is determined by a hyperplane through $x$ \cite{rousseeuw96RSS}. For example, the line through $x$ is rotated by $180^{\circ}$ in Figure \ref{fig:hd_bi}. The halfspace depth of $x$ is determined by the smallest portion of data that are separated by such a hyperplane (e.g.,  the halfspace depth of $x$ is $3/13$, determined by the dotted line).

@@

\subsection{OD Flow Outlier Detection Based on Depth}
The center-outward ordering in data depth is closely related to the detection of outliers.
The upper level sets of data depth in $\mathbb{R}^2$ form the central regions.
The most central region can be regarded as a median.
Conversely, the lower level sets of data depth, coincide with large distance from the center, can be regarded as outlyingness. \cite{rousseeuw99AS,aplpackR} utilizes this concept to generate a bag plot, which is analogous to the one-dimensional box plot based on data depth.
This paper uses the bag plot in order to identify the outliers of OD flows. 
Before explaining the method of outliers detection, we first introduce a couple of definitions.

@@

The bag plot is able to present the location (the depth median), spread (the size of bag), correlation (the orientation of the bag), and skewness (the shape of the bag and the fence) of data \cite{rousseeuw99AS}. For example, we can see which forward OD flows have higher counts compared to the reverse OD flows; the relatively linear correlation between forward OD flows and reverse OD flows; and the skewness of forward (reverse) OD flows in Figure \ref{fig:OD_0521}.

In addition, it is possible to detect the outliers of OD flows of two different time stamps. In Figure \ref{fig:OD_0721_0521}, we visualize the OD flows of two different days. Therefore, we can not only see where is the central region of OD flows, but also differentiate which OD flows are significantly different compared with two different OD flows.

Further, the OD flows of high active areas in a city are more likely to have large volume of trips.
We use set operations to detect such outliers. For example, we regard OD flows on July 1  as control data set ($control$); OD flows on May 21 as test data set ($test$); and the combination of two OD flows as combination data set ($combination$) in Figure \ref{fig:bagplot_0521_0701}. Then we can calculate the intersection of three outliers sets ($control \cap test \cap combination$), which are represented as rectangle symbols in Figure \ref{fig:OD_outliers_high}.

@@

\subsection{OD Flow Comparisons Based on Depth}
Data depth can compare bivariate data related with two independent groups \cite{liu93JASA}. 
A t-test can be used to compare means from two independent groups.
For example, the t-test reveals whether two OD flows' means are different at two different temporal ranges.
However, it is worth examining how groups differ in terms of scale.
The comparisons of central regions in data depth compare the marginal distribution, thereby considering the overall structure of the data \cite{wilcox03MBR}.
